# ggplot2

## Packages needed for these examples

* ggplot2 for the main plotting functions 
* forcats for easy factor releveling
* ggrepel for better in-plot labeling 
* ggpubr for easy signifigance bars and testing

```{r message=FALSE, results = "hide"}
library(ggplot2)
library(forcats)
library(ggrepel)
library(ggpubr)
```

## Boxplots

### Boxplot examples with iris data 

Setup the data in long format which is better suited for ggplot2.

```{r}
long_iris <- reshape(data = iris, direction = "long", 
                     timevar = "Dimension", 
                     times = colnames(iris)[1:4], 
                     idvar = "flower_ID", 
                     ids = paste0("f", 1:150), 
                     v.names = "Length", 
                     varying = 1:4, 
                     new.row.names = 1:600)
```


`forcats::fct_reorder()` used here to reorder factor levels for both Species (ascending left to right) and Sepal/Petal lengths.

```{r}
ggplot(long_iris, mapping = aes(x=forcats::fct_reorder(Species, Length), # reorder by measurements
                                y=Length, 
                                colour=forcats::fct_reorder(Dimension, Length))) + # reorder by measurements
  # Define fill as a alpha shade of colour aesthetic
  geom_boxplot(mapping = aes(fill = after_scale(alpha(colour, 0.4)))) +
  # Set colours manually 
  scale_color_manual(name="Dimension", values = c("dodgerblue", "red1", "dodgerblue4", "red4")) +
  # Set y-axis to end on tick, you could use c(0,0) but expansion offers more customization i.e. non-symmetrical expansion of axis
  scale_y_continuous(name = "Measurement (mm)", expand = expansion(mult = 0), limits = c(0,9), breaks = seq(0,9,3)) +
  theme_classic(13) +
  # Modify theme
  theme(title = element_text(face="bold"),
        legend.title = element_blank(),
        axis.text = element_text(colour="black"),
        axis.title.x = element_blank(),
        axis.ticks.length = unit(0.5, "lines"),
        panel.grid.major.y = element_line(colour = ggplot2::alpha("black", 0.4), linetype = "dashed"),
        legend.position = "inside", 
        legend.position.inside = c(0.12, 0.85),
        )
```


Sometimes we want to facet which enables a bit nicer X-axis options e.g. separating axis for different species. This also enables possibility to use `ggpubr::stat_compare_means` for testing.

```{r}
# Comparisons to use for testing
my_comparisons <- list( c("Petal.Width", "Sepal.Width"), c("Petal.Length", "Sepal.Length"))

# For testing we need to set the groups in the comparison on X-axis
ggplot(long_iris, mapping = aes(x=forcats::fct_reorder(Dimension, Length), 
                                y=Length, 
                                colour=forcats::fct_reorder(Dimension, Length))) +
  geom_boxplot(mapping = aes(fill = after_scale(alpha(colour, 0.4)))) +
  # P-value labeling using asterix 
  ggpubr::stat_compare_means(mapping = aes(label = after_stat(p.signif)), 
                             comparisons = my_comparisons, label.y = c(8,8)) +
  scale_color_manual(name="Dimension", values = c("dodgerblue", "red1", "dodgerblue4", "red4")) +
  scale_y_continuous(name = "Measurement (mm)", expand = expansion(mult = 0), limits = c(0,9), breaks = seq(0,9,3)) +
  # faceting and reordering the facets
  facet_wrap(~forcats::fct_reorder(Species, Length), strip.position="bottom") +
  theme_classic(13) +
  theme(title = element_text(face="bold"),
        axis.text = element_text(colour="black"),
        axis.title.x = element_blank(),
        axis.ticks.length = unit(0.5, "lines"),
        axis.ticks.x = element_blank(),
        strip.background = element_blank(),
        strip.placement = "outside",
        axis.text.x = element_blank(),
        panel.grid.major.y = element_line(colour = ggplot2::alpha("black", 0.4), linetype = "dashed")
        )
```

## Volcano plots 

### Volcano plot from the airway dataset

Some example volcano plots using airway dataset. First we have to perform de_analysis as the data is unprocessed.

```{r message=FALSE, results = "hide"}
library(airway)
library(DESeq2)
data("airway")

# Perform DEA
dds <- DESeqDataSet(airway, design = ~ cell + dex)
dds <- DESeq(dds, betaPrior=FALSE)
res <- results(dds, contrast = c('dex','trt','untrt'))
res <- lfcShrink(dds, contrast = c('dex','trt','untrt'), res=res, type = 'normal')

# Make data.frame for plotting
res_df <- as.data.frame(res)
res_df$ensid <- rownames(res_df) 
res_df$gene_name <- airway@rowRanges@elementMetadata@listData$gene_name[match(airway@rowRanges@elementMetadata@listData$gene_id, res_df$ensid)]

# DO NOT DO THIS WITH REAL DATA WITHOUT KNOWING WHAT YOU ARE ACTUALLY REMOVING HERE
res_df <- na.omit(res_df)
```




Simple volcano with labeling using `ggrepel::geom_text_repel()`.

```{r fig.width=4, fig.height=4}
ggplot(na.omit(res_df), mapping = aes(x=log2FoldChange, 
                                      y=-log10(padj), 
                                      colour=factor(ifelse(padj<0.01 & abs(log2FoldChange)>1, "Sig.", "NS"), 
                                                    levels = c("Sig.", "NS")))) +
  geom_point(alpha=0.8) +
  scale_color_manual(values = c(scales::muted("red"), ggplot2::alpha("grey80", 0.8))) +
  
  ggrepel::geom_text_repel(mapping = aes(label = ifelse(padj<1e-50 & abs(log2FoldChange)>2.5, gene_name, "")),
                          box.padding = 1, seed=1234, size=2, max.overlaps = Inf, min.segment.length = 0) +
  theme_bw() +
  # add cutoffs 
  geom_vline(xintercept = c(-1,1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.01), linetype = "dashed") +
  # Using bquote for subscripts
  scale_y_continuous(name = bquote(bold(-log[10](FDR))), 
                     expand = expansion(mult = 0), 
                     limits = c(0,150), 
                     breaks = seq(0,150,50)) +
  scale_x_continuous(name = bquote(bold(log[2](Fold~change))), 
                     expand = expansion(mult = 0), 
                     limits = c(-5,5), 
                     breaks = seq(-5,5,2.5)) +
  # prevent clipping of the panel borders
  coord_cartesian(clip = "off") +
  theme(axis.title = element_text(face="bold"),
        axis.text = element_text(colour="black"),
        axis.ticks.length = unit(0.5, "lines"),
        axis.line = element_blank(),
        panel.border = element_rect(colour = "black"),
        legend.title = element_blank(), 
        panel.grid = element_blank(),
        legend.position = "inside",
        legend.position.inside = c(0.1,0.9),
        legend.background = element_blank())
```



For a bit more customizable plots we can separate significant and non-significant DEGs

```{r}
# create dataframe for significant DEGs
res_df_sig <- res_df[res_df$padj<0.01 & abs(res_df$log2FoldChange)>1,]

# Create dataframe for non significant
res_df_nsig <- res_df[!(res_df$gene_name %in% res_df_sig$gene_name),]
```



More classic version with up and down regulated genes with different colours. 

```{r, fig.width=4, fig.height=4}
ggplot() +
  geom_point(data = res_df_nsig, mapping= aes(x=log2FoldChange, 
                                              y=-log10(padj)), 
             colour="grey80", alpha=0.4) +
  geom_point(data = res_df_sig, mapping= aes(x=log2FoldChange, 
                                             y=-log10(padj), 
                                             colour=factor(ifelse(log2FoldChange>0, "Up", "Down"), levels = c("Up", "Down"))), 
             alpha=0.8) +
  scale_colour_manual(values = c("darkred", "dodgerblue4")) +
  theme_bw() +
  geom_vline(xintercept = c(-1,1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.01), linetype = "dashed") +
  scale_y_continuous(name = bquote(bold(-log[10](FDR))), 
                     expand = expansion(mult = 0), 
                     limits = c(0,150), 
                     breaks = seq(0,150,50)) +
  scale_x_continuous(name = bquote(bold(log[2](Fold~change))), 
                     expand = expansion(mult = 0), 
                     limits = c(-5,5), 
                     breaks = seq(-5,5,2.5)) +
  # prevent clipping of the panel borders
  coord_cartesian(clip = "off") +
  theme(axis.title = element_text(face="bold"),
        axis.text = element_text(colour="black"),
        axis.ticks.length = unit(0.5, "lines"),
        axis.line = element_blank(),
        panel.border = element_rect(colour = "black"),
        legend.title = element_blank(), 
        panel.grid = element_blank(),
        legend.position = "none")
```


Some times it can be interesting to see how the DEGs are separated into up- and down-regulated ones. We can calculate densities in 2d space using `MASS` and map the values to points in the graph. 

```{r, fig.width=4, fig.height=4}
#For calculating density 
get_density <- function(x, y, ...) {
  dens <- MASS::kde2d(x, y, ...)
  ix <- findInterval(x, dens$x)
  iy <- findInterval(y, dens$y)
  ii <- cbind(ix, iy)
  return(dens$z[ii])
}

# calculate densities for points
res_df_sig$density <- get_density(-log10(res_df_sig$padj), res_df_sig$log2FoldChange)

# plot separately 
ggplot() +
  geom_point(data = res_df_nsig, mapping= aes(x=log2FoldChange, y=-log10(padj)), colour="grey80", alpha=0.4) +
  geom_point(data = res_df_sig, mapping= aes(x=log2FoldChange, y=-log10(padj), colour=density), alpha=0.8) +
  scale_color_viridis_c() +
  theme_bw() +
  geom_vline(xintercept = c(-1,1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.01), linetype = "dashed") +
  scale_y_continuous(name = bquote(bold(-log[10](FDR))), 
                     expand = expansion(mult = 0), 
                     limits = c(0,150), 
                     breaks = seq(0,150,50)) +
  scale_x_continuous(name = bquote(bold(log[2](Fold~change))), 
                     expand = expansion(mult = 0), 
                     limits = c(-5,5), 
                     breaks = seq(-5,5,2.5)) +
  # prevent clipping of the panel borders
  coord_cartesian(clip = "off") +
  theme(axis.title = element_text(face="bold"),
        axis.text = element_text(colour="black"),
        axis.ticks.length = unit(0.5, "lines"),
        axis.line = element_blank(),
        panel.border = element_rect(colour = "black"),
        legend.title = element_blank(), 
        panel.grid = element_blank(),
        legend.position = "none")
```


